---
title: Reproducible research - Project 1
author: "Shirshendu Nandy"
date: "18 July 2017"
output: html_document
---

The following analysis are performed using anonymised data collected during the months of October and November, 2012 that includes the number of steps taken in 5 minute intervals each day from a personal activity monitoring device.

Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)

The variables included in this dataset are:

* steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
* date: The date on which the measurement was taken in YYYY-MM-DD format
* interval: Identifier for the 5-minute interval in which measurement was taken

## 1 Loading and preprocessing the data
```{r}
#if(!file.exists("data")){dir.create("data")}
#fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
#download.file(fileUrl,destfile="./data/RepReserchProject1.zip")
#unzip(zipfile="./data/RepReserchProject1.zip",exdir="./data")
activity <- read.csv(".\\data\\activity.csv",header = T)
str(activity)
## As the date field is a factor variable, we will need to transform this into a date class.
as.Date(activity$date, "%Y-%m-%d") -> activity$date
```

## 2 What is mean total number of steps taken per day?
For this part of the assignment, you can ignore the missing values in the dataset.

**Calculate the total number of steps taken per day**

```{r}
StepsTotalPerDay <- aggregate(activity$steps, by=list(activity$date), FUN=sum, na.rm=T)
names(StepsTotalPerDay) <- c("Date", "Total")
head(StepsTotalPerDay)
```

**Make a histogram of the total number of steps taken each day**

```{r}
hist(StepsTotalPerDay$Total, col="grey", main="Histogram of total steps per day", xlab="Total steps")
```

**Calculate and report the mean and median of the total number of steps taken per day**

```{r}
summary(StepsTotalPerDay)
```

*  Mean and median are 9354 and 10395 respectively for total steps*

## 3 What is the average daily activity pattern?

**Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)**

```{r}
StepAveragesPerInterval <- aggregate(activity$steps, by=list(activity$interval), FUN = mean, na.rm = T)
names(StepAveragesPerInterval) <- c("Interval", "AverageSteps")
head(StepAveragesPerInterval)
plot(StepAveragesPerInterval, type="l", main="Time series of average number of steps taken by intervals", xlab="Interval", ylab="Frequency of average steps ")
```

**Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**
```{r}
StepAveragesPerInterval[which(StepAveragesPerInterval$AverageSteps == max(StepAveragesPerInterval$AverageSteps)), 1]
```

*Interval number 835 contains the maximum number of average steps* 

## Imputing missing values

**Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)**

```{r}
sum(is.na(activity$steps))
```

There are 2304 rows where the steps value are missing.


**Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.**

```{r}
# NA indices
NAIndex <- which(is.na(activity$steps))

# Create a vector of means
ImputeValues <- rep(mean(activity$steps, na.rm=TRUE), times=length(NAIndex))
```

**Modify the dataset with missing values replaced with a mean**
```{r}
activity[NAIndex, "steps"] <- ImputeValues
head(activity)
```

**Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?**

```{r}
StepsTotalPerDay <- aggregate(activity$steps, by=list(activity$date), FUN=sum, na.rm=T)
names(StepsTotalPerDay) <- c("Date", "Total")
hist(StepsTotalPerDay$Total, col="grey", main="Histogram of total steps per day (NAs removed)", xlab="Total steps")
summary(StepsTotalPerDay)
```

Both the mean and median for this modified set are 10,766 which is higher than previously observed values (Mean a: 9354 and median :10395)

## Are there differences in activity patterns between weekdays and weekends?

**Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day**

```{r}
# Calculate the weekdays 
activity <- cbind(activity, day= weekdays(activity$date))
## Calculate type of day ( weekend or week day)
activity  <- cbind(activity, daytype=ifelse(activity$day == "Saturday" | activity$day == "Sunday", "weekend", "weekday"))
head(activity)
```


**Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).**

```{r}
# Calculate average number of steps taken by daytype, day and interval
activityPlot <- aggregate(activity$steps, 
                       by=list(activity$daytype,activity$day, activity$interval), FUN = mean)

names(activityPlot) <- c("daytype", "day", "interval", "mean")
head(activityPlot)
```

```{r}
library(lattice)
xyplot(mean ~ interval | daytype, activityPlot, type="l", xlab="Interval", ylab="Number of steps",layout=c(1,2),  main="Average number of steps taken across weekdays or weekends")
```
